---
title: "Programming Assignment 1 - Reproducible Research"
author: "Sangeet Khullar"
date: "May 17, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read CSV Input

```{r readData}
#setwd("C:\\Users\\skhullar\\OneDrive - Daiichi Sankyo\\Documents\\GitHub\\Reproducible_Research")
act <- read.csv("activity.csv")
dim(act)
```

## Find Mean & Median Steps Per Day   

**1. Sum up all the steps for a given Day - ( Sum(steps) - group by Date )**


**2. Display Histogram of Steps Per Day** 


```{r histogramStepsPerDay, echo=TRUE }
library(dplyr)
steps_per_day_dframe <- act    %>% group_by(date) %>% summarize(steps_per_day=sum(steps))   
hist(steps_per_day_dframe$steps_per_day)
```


**3. Display BarPlot of Steps Per Day**


```{r barPlotStepsPerDay, echo=TRUE }
barplot(steps_per_day_dframe$steps_per_day)
```

**4. Find Mean & Median value per day  from the above aggregated dataset**  

```{r findMeanNmedianStepsPerDay, echo=TRUE }
mean_steps_per_day <- steps_per_day_dframe %>% summarize(mean_steps_per_day=mean(steps_per_day,na.rm = TRUE))
print(paste("Mean_Steps_Per_Day=",mean_steps_per_day))



median_steps_per_day <- steps_per_day_dframe %>% summarize(median_steps_per_day=median(steps_per_day,na.rm = TRUE))
print(paste("Median_Steps_Per_Day=",median_steps_per_day))

```


## Average daily activity pattern

**1. Time series plot of the 5-minute interval & the average number of steps taken, averaged across all days** 


```{r timeSeriesPlot_DailyActivity, echo=TRUE }
library(ggplot2)

avg_steps_per_5min_interval <- act    %>% group_by(interval) %>% summarize(avg_steps_for_interval= mean(steps,na.rm = TRUE))
  
ggplot(data = avg_steps_per_5min_interval, aes(x = interval, y = avg_steps_for_interval))+
  geom_line(color = "#00AFBB", size = 2) + 
   geom_smooth(method = "lm") + 
     labs(x = "Time Interval in Seconds ") +
          labs(y = "Average Number of Steps") + 
            labs(title = "Time Series Data for Number of Steps Across the Day")



```


**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**


```{r findMaxStepAverageInterval, echo=TRUE }
max_step_average <- max(avg_steps_per_5min_interval$avg_steps_for_interval, na.rm = TRUE)

max_step_average_interval <- avg_steps_per_5min_interval[which(avg_steps_per_5min_interval$avg_steps_for_interval == 
                                   max_step_average), ]
print(paste("Max_Step_Average_Interval=",max_step_average_interval$interval))
```



## Imputing missing values

**Report the total number of missing values in the dataset**
```{r countNA, echo=TRUE }
summary(act)
count_na_values <- sum(is.na(act$steps))
print(paste("Count_Of_Records_With_Missing_Step_Values=",count_na_values))
```


**Fill up the missing values in the dataset** 
```{r imputeDataFrame, echo=TRUE }
library(mice)
tempData <- mice(act,m=5,maxit=50,meth='pmm',seed=500)
completedData <- complete(tempData,1)
```



**Create a new Data Set with missing values populated**
```{r newImputeDataFrame, echo=TRUE }
completedData <- complete(tempData,1)
```

**Make a histogram of the total number of steps taken each day** 
```{r histogramStepsPerDayAfterImput, echo=TRUE }
steps_per_day_dframe_after_imput <- completedData    %>% group_by(date) %>% summarize(steps_per_day=sum(steps))   
hist(steps_per_day_dframe_after_imput$steps_per_day)
```



**Calculate the mean and median total number of steps taken per day AFTER Imput** 
```{r findMeanNmedianStepsPerDayAfterImput, echo=TRUE }
mean_steps_per_day_after_imput  <- steps_per_day_dframe_after_imput %>% summarize(mean_steps_per_day=mean(steps_per_day,na.rm = TRUE))
print(paste("Mean_Steps_Per_Day_After_Imput=",mean_steps_per_day_after_imput))



median_steps_per_day_after_imput <- steps_per_day_dframe_after_imput %>% summarize(median_steps_per_day=median(steps_per_day,na.rm = TRUE))
print(paste("Median_Steps_Per_Day_After_Imput=",median_steps_per_day_after_imput))

```


**Mean & Median Before Imput** 
```{r MeanNmedianStepsPerDayBeforeImput, echo=TRUE }
print(paste("Mean_Steps_Per_Day_Before_Imput=",mean_steps_per_day))

print(paste("Median_Steps_Per_Day_Before_Imput=",median_steps_per_day))

```

**Impact of imputing missing data on the estimates of the total daily number of steps**
```{r impactOfImputing, echo=TRUE }
diff_mean <-  mean_steps_per_day_after_imput -mean_steps_per_day
print(paste("Change_in_Mean_Steps_Per_day=",diff_mean))

diff_median <- median_steps_per_day_after_imput - median_steps_per_day
print(paste("Change_in_Median_Steps_Per_day=",diff_median))

```


## Differences in activity patterns between weekdays and weekends

**New factor variable weekday and weekend**
```{r weekDayNweekEnd, echo=TRUE }
for (row in 1:nrow(completedData)) {
    day  <- weekdays(as.Date(completedData[row, "date"]))
       if(day %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")) 
         {
        completedData[row,"is_Weekday"] <- "1" 
         }
        else 
          {
        completedData[row,"is_Weekday"] <- "0"    
          }
}
```


**Panel Time Series Plot time series plot** 
```{r timeSeries_WeekDayNWeekEnd, echo=TRUE }
avg_steps_per_5min_interval_imp <- completedData  %>% group_by(interval,is_Weekday) %>% 
      summarize(avg_steps_for_interval= mean(steps,na.rm = TRUE))
  
ggplot(data = avg_steps_per_5min_interval_imp, aes(x = interval, y = avg_steps_for_interval ,color=is_Weekday))+
   facet_grid( .~ is_Weekday) + 
    geom_smooth(method = "lm") + 
     labs(x = "Time Interval in Seconds ") +
          labs(y = "Average Number of Steps") + 
            labs(title = "Time Series Data for Number of Steps Across the Day")
```


## The End 